Versioned Queries over RDF Archives:\\All You Need is SPARQL?

### 2020-10-09 paper.tex
camera ready version

\usepackage{floatrow}
% Table float box with bottom caption, box width adjusted to content
\newfloatcommand{capbtabbox}{table}[][\FBwidth]

@@

\begin{abstract}
We explore solutions for representing archives of versioned RDF data using the SPARQL standard and off-the-shelf engines. We consider six representations of RDF archives based on named graphs, and describe how input queries can be rewritten to return solutions for a particular version, or solutions that change between versions. We evaluate these alternatives over an archive of 8 weekly versions of Wikidata and 146 queries, using Virtuoso as the SPARQL engine.
\end{abstract}

@@

A key aspect of the Web is its dynamic nature, where documents are frequently updated, deleted and added. Likewise when we speak of the Semantic Web, it is important to consider that sources may be dynamic and RDF datasets are subject to change~\cite{KaferAUOH13}. It is in this context that various works have looked at versioning in the context of RDF/SPARQL~\cite{VolkelG06,TappoletB09,Grandi10,GraubeHU14,KhuranaD16}, with recent works proposing \textit{RDF archives}~\cite{FernandezPU15,Cerdeira-PenaFF16,BahriLA18,FernandezUPK19,TaelmanSHMV19} that manage RDF graphs and their historical changes, allowing for querying across different versions of the graph. Within these works, a variety of specialised indexing techniques~\cite{Cerdeira-PenaFF16,BahriLA18,TaelmanSHMV19}, query languages~\cite{TappoletB09} and benchmarks~\cite{KotsevMPEFK16,FernandezUPK19} have been proposed, developed and evaluated. While these represent important advances, many such works propose custom SPARQL extensions, engines, etc., creating an barrier for adoption.

@@

In fact, the types of versioned queries proposed in the literature~\cite{FernandezPU15} can be supported using off-the-shelf SPARQL engines that already enjoy years of development, optimisation, and deployment. SPARQL named graphs can, for exampe, be used to track different versions of individual graphs. However, as Fernandez at al.~\cite{FernandezUPK19} note, the approach of using pure SPARQL would ``\textit{typically render rather inefficient SPARQL queries}''. This raises a question: how inefficient will such queries be? If a pure SPARQL solution could be found with reasonable performance, existing SPARQL infrastructure could be leveraged for RDF archives.

@@

More recently the concept of RDF archives have been gaining more and more attention. Fernandez et al.~\cite{FernandezPU15} provide a general survey and introduction to the theme, discussing the types of queries that one might wish to answer in the context of an RDF archive. Works by Cerdeira{-}Pena et al.~\cite{Cerdeira-PenaFF16}, Zaniolo et al.~\cite{ZanioloGACG18} and Taelman et al.~\cite{TaelmanSHMV19} propose ways to compress and index RDF archives, while Khurana and Deshpande~\cite{KhuranaD16} propose indexes more generally for historical graph data. Bahri et al.~\cite{BahriLA18} explore the benefits of using Apache Spark to manage RDF archives in a distributed setting. In order to evaluate these emerging techniques, a number of benchmarks have been proposed, including the BEnchmark of RDF
ARchives (BEAR)~\cite{FernandezUPK19}, and the Semantic Publishing Benchmark (SPB)~\cite{Papakonstantinou18}.

The past years have thus seen a wide range of developments in terms of representations, languages, vocabularies, indexes, algorithms, etc., for managing and querying over temporal and/or versioned RDF data. However, to the best of our knowledge, most of these approaches propose specialised languages, implementations, etc., for these purposes, creating an obstacle for adoption. %Such languages have the benefit of allowing to more concisely and intuitively express versioned queries, while such implementations can develop optimised techniques that leverage particular characteristics of versioned data. The downside of these specialised approaches is that they are often diverge from practice, requiring the use of research prototypes rather than established SPARQL implementations, non-standard languages, etc. 
A number of authors note that one can manage and query RDF archives using vanilla SPARQL, though it may lead to prolix or inefficient queries~\cite{TappoletB09,FernandezUPK19}. This paper describes preliminary experiments to gain insights into how (in)efficient vanilla SPARQL and off-the-shelf engines are for hosting RDF archives.

@@

We establish some preliminaries and notation. RDF triples are composed of three sets of terms: IRIs (\I), literals (\L) and blank nodes (\B). In this preliminary work, we do not consider blank nodes, whose special semantics complicates the detection of changes~\cite{ZeginisTC11}. Leaving aside blank nodes, an RDF triple $(s,p,o) \in \I \times \I \times (\I \cup \L)$ consists of a subject $s$, predicate $p$ and object $o$. An RDF graph $G$ is then a set of triples. An RDF archive is a tuple of RDF graphs $\mathcal{G} = (G_1,\ldots,G_n)$, interpreted as $n$ sequential versions of a given RDF graph.

@@

\begin{table}[t]
{\footnotesize
\begin{align*}
\sand{P}{Q}(G) \da &\, P(G) \Join Q(G) & M_1 \Join M_2 \da &\, \{ \mu_1 \cup \mu_2 \mid \mu_1 \in M_1,\, \mu_2 \in M_2,\, \com{\mu_1}{\mu_2} \} \\
\suni{P}{Q}(G) \da &\, P(G) \cup Q(G) & M_1 \cup M_2 \da &\, \{ \mu \mid \mu \in M_1\text{ or }\mu \in M_2 \} \\
\sminus{P}{Q}(G) \da &\, P(G) \setminus Q(G) &  M_1 \setminus M_2 \da &\, \{ \mu_1 \in M_1 \mid \nexists \mu_2 \in M_2 : \com{\mu_1}{\mu_2} \} \\
%M_1 \loj M_2 \da\, & (M_1 \bowtie M_2) \cup (M_1 \setminus M_2) \\
%\sigma_R(M) \da\,& \{ \mu \in M \mid \mu \models R \}\\
%\ssel{P}{V}(G) \da &\, \pi_V(P(G)) & \pi_V(M) \da\,& \{ \mu' \mid \exists \mu \in M : \mu' \sim \mu, \dom{\mu'} = V \cap \dom{\mu} \}
\end{align*}}
\vspace{-1cm}
\caption{Semantics for query operators where $M$, $M_1$ and $M_2$ denote sets of solutions\label{tab:sparql}}
\end{table}

%A basic graph pattern $B$ is a set of triple patterns. %Let $\vars{B}$ denote the set of variables appearing in $B$ and let $\mu(B)$ denote the image of $B$ under $\mu$, i.e., the result of replacing all variables $v \in \dom{\mu} \cap \vars{B}$ with $\mu(v)$ in $B$. The evaluation of a basic graph pattern $B$ on an RDF graph $G$ is given as $B(G) \coloneqq \{ \mu \mid \mu(B) \subseteq G\text{ and }\dom{\mu} = \vars{B} \}$. 
SPARQL queries are based on triple patterns and a number of relational operators. Similar to Pérez et al.~\cite{PerezAG09}, we define an abstract syntax for a subset of SPARQL of pertinence to this paper as follows. A triple pattern $t$ is a graph pattern. Furthermore, if $P$ and $Q$ are graph patterns, and $V \subset \V$ is a set of variables, then $\sand{P}{Q}$, $\suni{P}{Q}$ and $\sminus{P}{Q}$ are graph patterns. We define the semantics of these graph patterns in Table~\ref{tab:sparql}.\footnote{A basic graph pattern $\{ t_1, \ldots t_n\}$ can be represented as $\sand{\sand{t_1}{\ldots}}{t_n}$.}

@@

\paragraph{Change-Based (CB):} The core idea of CB representations is to store only triples that change from a given reference version. Along these lines, in the following we denote by $\Delta_{i}^j \coloneqq G_i \setminus G_j$ the triples in version $i$ not in version $j$. We consider four CB representations based on four transformations of $\mathcal{G} = (G_1,\ldots,G_n)$:
%
{\footnotesize
\begin{align*}
\mathcal{G}^n_1 \da\,& (G_1,\Delta_1^2,\Delta_2^1,\ldots,\Delta_1^n,\Delta_n^1) & G_i & = (G_1 \cup \Delta_i^1) \setminus \Delta_1^i & \text{for }1 < i \leq n\\
\mathcal{G}^n_{n-1} \da\,& (G_1,\Delta_1^2,\Delta_2^1,\ldots,\Delta_{n-1}^n,\Delta_n^{n-1}) & G_i & = (G_{i-1} \cup \Delta_i^{i-1}) \setminus \Delta_{i-1}^i & \text{for }1 < i \leq n \\
\mathcal{G}^1_n \da\,& (G_n,\Delta_n^{n-1},\Delta_{n-1}^n,\ldots,\Delta_n^1,\Delta_1^n) & G_i & = (G_n \cup \Delta_i^n) \setminus \Delta_n^i & \text{for }1 \leq i < n \\
\mathcal{G}^{n-1}_n \da\,& (G_n,\Delta_n^{n-1},\Delta_{n-1}^n,\ldots,\Delta_2^1,\Delta_1^2) & G_i & = (G_{i+1} \cup \Delta_i^{i+1}) \setminus \Delta_{i+1}^i & \text{for }1 \leq i < n
\end{align*}%
}%
%
As can be seen for the definitions of $G_i$, these transformations are lossless: we can retrieve any version of the graph from any such transformation. The first two transformations start with the earliest version as a base. The first encodes deltas always with respect to the first version. The second encodes deltas with respect to the previous version. The latter two transformations start with the latest version. The third encodes deltas with respect to the latest version. The forth encodes deltas with respect to the subsequent version. Letting $\mathcal{H} = (H_1,\ldots,H_n)$ such that $H_i = G_{n-i+1}$ ($1 \leq i \leq n$), i.e., such that $\mathcal{H}$ ``reverses'' $\mathcal{G}$, we remark that $\mathcal{G}^n_1 = \mathcal{H}^1_n$ and $\mathcal{G}^n_{n-1} = \mathcal{H}^{n-1}_n$. Each such transformation can then be represented as a SPARQL dataset with $2(n-1)+1$ named graphs.

In terms of space we expect $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$ to be the most efficient as they always encode deltas from a neighbouring version. However, in terms of query rewriting, we expect $\mathcal{G}^n_{1}$ and $\mathcal{G}^{1}_n$ to be more efficient as they do not require a recursive construction of all intermediate versions towards the base version. In terms of indexing a new version, we expect $\mathcal{G}^{n}_1$ followed by $\mathcal{G}^{n}_{n-1}$ to be most efficient as they require only computing the most recent deltas; we expect $\mathcal{G}^{1}_n$, followed by $\mathcal{G}^{n-1}_{n}$, to be much more expensive, requiring a recompute of all deltas. On the other hand, $\mathcal{G}^{1}_n$ and $\mathcal{G}^{n-1}_{n}$ should be advantageous for queries over more recent versions, which we could reasonably expect to be the most frequent.

@@

\paragraph{Timestamp-Based (TB):} The intuition of the TB representation is to associate each triple with the versions in which it is contained. Along these lines, we denote by $\mathcal{G}(s,p,o) \coloneqq \{ i \mid (s,p,o) \in G_i\text{ for } 1 \leq i \leq n \}$ the versions containing $(s,p,o)$.\footnote{The definition $\mathcal{G} : \I \times \I \times (\I \cup \L) \rightarrow 2^{\{1,\ldots,n\}}$~\cite{FernandezUPK19} is analogous to $\mathcal{G} = (G_1,\ldots,G_n)$.} Let $\mathcal{N} \coloneqq \{ N \mid \exists (s,p,o) \in G_i : \mathcal{G}(s,p,o) = N, 1 \leq i \leq n  \}$ denote the family of sets of versions associated with some triple in $(s,p,o)$. We can then represent the RDF archive with a named graph for each $N \in \mathcal{N}$. However, care must be taken as the number of named graphs could potentially reach $2^n$ (or the total number of unique triples in $\mathcal{G}$). A more practical option is create a named graph for intervals~\cite{FernandezPU15}. More specifically, a triple $(s,p,o)$ is added to an interval $[i,j]$ (for $1 \leq i \leq j \leq n$) if and only if $(s,p,o) \in G_k$ for $i \leq k \leq j$ and either $i = 1$ or $(s,p,o) \notin G_{i-1}$ and $j = n$ or $(s,p,o) \notin G_{j+1}$; in simpler terms, $[i,j]$ is a maximal contiguous interval of versions in which $(s,p,o)$ appears (omitting empty intervals). The upper bound is now $n(n+1)/2$ named graphs, but with the possibility of triples appearing in multiple named graphs for different intervals.

In general, we expect the space to be similar to $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$; in other words, quite good. In terms of query rewriting, we would expect performance to be quite good for a low number of versions and few intermittent triples as -- unlike for CB representations -- negation is not required; we can rather take the union of interval graphs containing the version. However, as the number of versions $n$ grows, $O(n^2)$ interval graphs need to be unioned in the worst-case to materialise a particular version; by way of comparison, CB representations require processing $O(1)$ (in the case of $\mathcal{G}^n_1$ and $\mathcal{G}^1_n$) or $O(n)$ (in the case of $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$) named graphs to materialise a particular version.

\paragraph{Notation:} We denote IC by $\ic$; incremental, differential, reverse-incremental, reverse-differential CB by $\cbpd$, $\cbpi$, $\cbmd$, $\cbmi$, resp.; and interval TB by $\tb$.

@@

Given a SPARQL query $Q$ defined for an RDF graph $G$, we now describe rewritings of $Q$ to generate solutions for different versions. We first focus on rewritings of triple patterns and then generalise. We do not extend the SPARQL syntax, but rather assume that version parameters are defined in (e.g.) the HTTP request.

@@

\paragraph{Change-Based (CB)} Recalling the observation that forwards and reverse CB representations are analogous, for brevity we define the rewriting for the forwards direction (\cbpd, \cbpi) only; the reverse direction (\cbmd, \cbmi) follows naturally. For the differential representation, we load the base version and the positive delta into the default graph and, for each triple pattern, we subtract the negative delta which is queried using a quad pattern. Thus for $\cbpd$ and $\cbmd$ we rewrite $Q$ to  \sfrom{Q'}{\{\texttt{1},\texttt{v1}\}}{\{\texttt{1v}\}}, where \texttt{1}, \texttt{v1}, \texttt{1v} indicate the names of $G_1$, $\Delta_v^1$ and $\Delta_1^v$, respectively; and $Q'$ replaces each triple pattern $(s,p,o) \in Q$ with the named graph pattern $\sminus{(s,p,o,*)}{(s,p,o,\texttt{1v})}$.

Unfortunately the incremental rewriting is more complex. A first idea would be to take the union of $G_1$ and all positive deltas $\Delta_2^1, \ldots, \Delta_v^{v-1}$ and then subtract the (union of the) negative deltas $\Delta_1^2, \ldots, \Delta_{v-1}^{v}$; unfortunately this would overlook triples that were removed from a version $1 \leq u < v$ but were added back in a later version $u < u' \leq v$ (and were not removed again in a version $u' < u'' \leq v$). Hence a recursive rewriting appears to be necessary. Let $Q_1 \coloneqq \sfrom{Q}{\{\texttt{1}\}}{\{\}}$; then $Q_2 \coloneqq \sfrom{Q'_1}{\{\texttt{1}\}}{\{\texttt{12},\texttt{21}\}}$, where $Q_1$ is the result of replacing each triple pattern $(s,p,o)$ in $Q_1$ by the named graph pattern $P_2 \coloneqq \sminus{\suni{(s,p,o,*)}{(s,p,o,\texttt{21})}}{(s,p,o,\texttt{12})}$. We can then apply this rewriting recursively: $Q_i \coloneqq \sfrom{Q_{i-1}'}{\{\texttt{1}\}}{\{\texttt{12},\ldots,\texttt{(i-1)i},\texttt{21},\ldots,\texttt{i(i-1)}\}}$, where $Q_{i-1}'$ replaces each named graph pattern $P_{i-1}$ appearing in $Q_{i-1}$ with the recursive pattern $P_i \coloneqq \sminus{\suni{P_{i-1}}{(s,p,o,\texttt{i(i-1)})}}{(s,p,o,\texttt{(i-1)i})}$.

This rewriting leads to complex queries. We thus optimise using additional features of SPARQL. To sketch the idea, our goal is then to make sure that each triple pattern only matches triples that appear in a base version and were not removed, or that appear in a positive delta $\Delta_{j}^{i}$ such that $1 < i < j \leq v$ and do not appear in a (later) negative delta $\Delta_{k}^{l}$ for $j \leq k < l \leq v$. In practice, for each delta $\Delta_{j}^{i}$ stored as a named graph $\texttt{ji}$, in offline processing, we index meta-data of the form $(\texttt{ji},\texttt{ver},j,\texttt{\$})$, and $(\texttt{ji},\texttt{type},\texttt{pos},\texttt{\$})$ in the case that $i < j$ or $(\texttt{ij},\texttt{type},\texttt{neg},\texttt{\$})$ in the case that $j < i$ ($\texttt{\$} \in \I$ is a reserved name for the meta-data graph). We can then check the aforementioned condition by using aggregation to find the maximum version of a negative delta less than or equals $v$ in which the triple pattern matches, then filtering the base version or any positive delta earlier than this maximum version. While the resulting query is still quite complex, we found it to be more practical than the recursive rewriting.

@@

Given a query $Q$, a control version $u$ and a target version $v$, delta-version queries give solutions in $Q(G_v) \setminus Q(G_u)$. The general strategy for rewriting is to construct a query $\sminus{Q_v}{Q_u}$ where $Q_v$ and $Q_u$ are the respective single-version queries. While we continue to use the default graph as before for $Q_v$, we must rather use named graphs to manage the positive elements of $Q_u$ (replacing \texttt{FROM} with \texttt{FROM NAMED}/\texttt{GRAPH}). As an example, for $\ic$, we rewrite $Q$ to \sfrom{Q'}{\{\texttt{v}\}}{\{\texttt{u}\}}, where $Q'$ replaces each triple pattern $(s,p,o)$ in $Q$ with $\sminus{(s,p,o,*)}{(s,p,o,\texttt{u})}$.

@@

\paragraph{Indexing} We first look at the results of total index sizes for each representation. In Figure~\ref{fig:index} we show the index sizes (GB) for each representation, the time taken (min.) to bulk load all versions in the representation, and the time taken (min.) to update a seven-version archive with the eighth version. We see that \ic has the largest index sizes, followed by \cbpd and \cbmd, then \tb, and finally \cbpi and \cbmi. The bulk load times correlate with index size, with \ic (thus) being by far the slowest. We see a similar trend in version updates, except that \cbmi is far slower than the other alternatives (even \ic) as the entire archive must be built from scratch.

>>>>

%\usepackage{floatrow}
% Table float box with bottom caption, box width adjusted to content
%\newfloatcommand{capbtabbox}{table}[][\FBwidth]

@@

\begin{abstract}
We explore solutions for representing archives of versioned RDF data using the SPARQL standard and off-the-shelf engines. We consider six representations of RDF archives based on named graphs, and describe how input queries can be automatically rewritten to return solutions for a particular version, or solutions that change between versions. We evaluate these alternatives over an archive of 8 weekly versions of Wikidata and 146 queries using Virtuoso as the SPARQL engine.
\end{abstract}

@@

A key aspect of the Web is its dynamic nature, where documents are frequently updated, deleted and added. Likewise when we speak of the Semantic Web, it is important to consider that sources may be dynamic and RDF datasets are subject to change~\cite{KaferAUOH13}. It is in this context that various works have looked at versioning in the context of RDF/SPARQL~\cite{VolkelG06,TappoletB09,Grandi10,GraubeHU14,KhuranaD16}, with recent works proposing \textit{RDF archives}~\cite{FernandezPU15,Cerdeira-PenaFF16,BahriLA18,FernandezUPK19,TaelmanSHMV19} that manage RDF graphs and their historical changes, allowing for querying across different versions of the graph. Within these works, a variety of specialised indexing techniques~\cite{Cerdeira-PenaFF16,BahriLA18,TaelmanSHMV19}, query languages~\cite{TappoletB09} and benchmarks~\cite{KotsevMPEFK16,FernandezUPK19} have been proposed, developed and evaluated. While these represent important advances, many such works propose custom SPARQL extensions, indexes, engines, etc., creating a barrier for adoption.

@@

In fact, versioned queries as proposed in the literature~\cite{FernandezPU15} can be supported using off-the-shelf SPARQL engines with years of development, optimisation, and deployment. SPARQL named graphs can, for example, be used to track different versions of individual graphs. However, as Fernandez at al.~\cite{FernandezUPK19} note, the approach of using pure SPARQL would ``\textit{typically render rather inefficient SPARQL queries}''. This raises a question: how inefficient will such queries be? If a pure SPARQL solution could be found with reasonable performance, existing SPARQL engines could be used to host and query RDF archives.

@@

More recently \textit{RDF archives} (of historical RDF data) have been gaining attention. Fernandez et al.~\cite{FernandezPU15} survey the theme, discussing the types of queries that can be run on such archives. Cerdeira{-}Pena et al.~\cite{Cerdeira-PenaFF16}, Zaniolo et al.~\cite{ZanioloGACG18} and Taelman et al.~\cite{TaelmanSHMV19} propose compressed indexes for RDF archives, while Khurana and Deshpande~\cite{KhuranaD16} propose indexes for historical graph data. Bahri et al.~\cite{BahriLA18} use Apache Spark to manage RDF archives in a distributed setting. Benchmarks have also been proposed for RDF archives, including the BEnchmark of RDF
ARchives (BEAR)~\cite{FernandezUPK19}, and the Semantic Publishing Benchmark (SPB)~\cite{Papakonstantinou18}.

The past years have seen many developments for managing and querying temporal, versioned or historical RDF data. But most of these approaches propose specialised languages, implementations, etc., creating an obstacle for adoption. A number of authors note that one can manage and query RDF archives using vanilla SPARQL, though it may lead to complex or inefficient queries~\cite{TappoletB09,FernandezUPK19}. Recently SPARQL has been used to host and query the edit history of Wikidata, but only one representation is explored~\cite{TanonS19}. This paper describes preliminary experiments to gain insights into the efficiency of off-the-shelf SPARQL engines for hosting RDF archives using different types of representations.

@@

RDF triples are composed of three sets of terms: IRIs (\I), literals (\L) and blank nodes (\B). We do not consider blank nodes in this work as they complicate the detection of changes~\cite{ZeginisTC11}. An RDF triple $(s,p,o) \in \I \times \I \times (\I \cup \L)$ consists of a subject $s$, predicate $p$ and object $o$. An RDF graph $G$ is a set of triples. An RDF archive is a tuple of RDF graphs $\mathcal{G} = (G_1,\ldots,G_n)$.

@@

\begin{figure}[t]
{\footnotesize
\begin{align*}
\sand{P}{Q}(G) \da &\, P(G) \Join Q(G) & M_1 \Join M_2 \da &\, \{ \mu_1 \cup \mu_2 \mid \mu_1 \in M_1,\, \mu_2 \in M_2,\, \com{\mu_1}{\mu_2} \} \\
\suni{P}{Q}(G) \da &\, P(G) \cup Q(G) & M_1 \cup M_2 \da &\, \{ \mu \mid \mu \in M_1\text{ or }\mu \in M_2 \} \\
\sminus{P}{Q}(G) \da &\, P(G) \setminus Q(G) &  M_1 \setminus M_2 \da &\, \{ \mu_1 \in M_1 \mid \nexists \mu_2 \in M_2 : \com{\mu_1}{\mu_2} \} \\
%M_1 \loj M_2 \da\, & (M_1 \bowtie M_2) \cup (M_1 \setminus M_2) \\
%\sigma_R(M) \da\,& \{ \mu \in M \mid \mu \models R \}\\
%\ssel{P}{V}(G) \da &\, \pi_V(P(G)) & \pi_V(M) \da\,& \{ \mu' \mid \exists \mu \in M : \mu' \sim \mu, \dom{\mu'} = V \cap \dom{\mu} \}
\end{align*}}
\vspace{-1cm}
\caption{Semantics for query operators where $M$, $M_1$ and $M_2$ denote sets of solutions\label{tab:sparql}}
\end{figure}

%A basic graph pattern $B$ is a set of triple patterns. %Let $\vars{B}$ denote the set of variables appearing in $B$ and let $\mu(B)$ denote the image of $B$ under $\mu$, i.e., the result of replacing all variables $v \in \dom{\mu} \cap \vars{B}$ with $\mu(v)$ in $B$. The evaluation of a basic graph pattern $B$ on an RDF graph $G$ is given as $B(G) \coloneqq \{ \mu \mid \mu(B) \subseteq G\text{ and }\dom{\mu} = \vars{B} \}$. 
SPARQL queries are based on triple patterns and a number of relational operators. Similar to Pérez et al.~\cite{PerezAG09}, we define an abstract syntax for a subset of SPARQL of pertinence to this paper as follows. A triple pattern $t$ is a graph pattern. Furthermore, if $P$ and $Q$ are graph patterns, and $V \subset \V$ is a set of variables, then $\sand{P}{Q}$, $\suni{P}{Q}$ and $\sminus{P}{Q}$ are graph patterns. We define the semantics of these graph patterns in Figure~\ref{tab:sparql}.\footnote{A basic graph pattern $\{ t_1, \ldots t_n\}$ can be represented as $\sand{\sand{t_1}{\ldots}}{t_n}$.}

@@

\paragraph{Change-Based (CB):} The core idea of CB representations is to store only triples that change from a given reference version. Along these lines, in the following we denote by $\Delta_{i}^j \coloneqq G_i \setminus G_j$ the triples in version $i$ not in version $j$. We consider four CB representations based on four transformations of $\mathcal{G} = (G_1,\ldots,G_n)$:
%
{\footnotesize
\begin{align*}
\mathcal{G}^n_1 \da\,& (G_1,\Delta_1^2,\Delta_2^1,\ldots,\Delta_1^n,\Delta_n^1) & G_i & = (G_1 \cup \Delta_i^1) \setminus \Delta_1^i & \text{for }1 < i \leq n\\
\mathcal{G}^n_{n-1} \da\,& (G_1,\Delta_1^2,\Delta_2^1,\ldots,\Delta_{n-1}^n,\Delta_n^{n-1}) & G_i & = (G_{i-1} \cup \Delta_i^{i-1}) \setminus \Delta_{i-1}^i & \text{for }1 < i \leq n \\
\mathcal{G}^1_n \da\,& (G_n,\Delta_n^{n-1},\Delta_{n-1}^n,\ldots,\Delta_n^1,\Delta_1^n) & G_i & = (G_n \cup \Delta_i^n) \setminus \Delta_n^i & \text{for }1 \leq i < n \\
\mathcal{G}^{n-1}_n \da\,& (G_n,\Delta_n^{n-1},\Delta_{n-1}^n,\ldots,\Delta_2^1,\Delta_1^2) & G_i & = (G_{i+1} \cup \Delta_i^{i+1}) \setminus \Delta_{i+1}^i & \text{for }1 \leq i < n
\end{align*}%
}%
%
As can be seen for the definitions of $G_i$, these transformations are lossless: we can retrieve any version of the graph from any such transformation. The first two transformations start with the earliest version as a base. The first encodes deltas always with respect to the first version. The second encodes deltas with respect to the previous version. The latter two transformations start with the latest version. The third encodes deltas with respect to the latest version. The forth encodes deltas with respect to the subsequent version. Letting $\mathcal{H} = (H_1,\ldots,H_n)$ such that $H_i = G_{n-i+1}$ ($1 \leq i \leq n$), i.e., such that $\mathcal{H}$ ``reverses'' $\mathcal{G}$, we remark that $\mathcal{G}^n_1 = \mathcal{H}^1_n$ and $\mathcal{G}^n_{n-1} = \mathcal{H}^{n-1}_n$. Each such transformation can then be represented as a SPARQL dataset with $2n - 1$ named graphs.

In terms of space we expect $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$ to be the most efficient as they always encode deltas from a neighbouring version. However, in terms of query rewriting, we expect $\mathcal{G}^n_{1}$ and $\mathcal{G}^{1}_n$ to be more efficient as they do not require a recursive construction of all intermediate versions towards the base version. In terms of indexing a new version, we expect $\mathcal{G}^{n}_1$ followed by $\mathcal{G}^{n}_{n-1}$ to be most efficient as they require only computing the most recent deltas; we expect $\mathcal{G}^{1}_n$, followed by $\mathcal{G}^{n-1}_{n}$, to be much more expensive, requiring a recompute of all deltas. On the other hand, $\mathcal{G}^{1}_n$ and $\mathcal{G}^{n-1}_{n}$ should be advantageous for queries over more recent versions, and in particular over the most recent version (a common case).

@@

\paragraph{Timestamp-Based (TB):} The intuition of the TB representation is to associate each triple with the versions in which it is contained. Along these lines, we denote by $\mathcal{G}(s,p,o) \coloneqq \{ i \mid (s,p,o) \in G_i\text{ for } 1 \leq i \leq n \}$ the versions containing $(s,p,o)$.\footnote{The definition $\mathcal{G} : \I \times \I \times (\I \cup \L) \rightarrow 2^{\{1,\ldots,n\}}$~\cite{FernandezUPK19} is analogous to $\mathcal{G} = (G_1,\ldots,G_n)$.} Let $\mathcal{N} \coloneqq \{ N \mid \exists (s,p,o) \in G_i : \mathcal{G}(s,p,o) = N, 1 \leq i \leq n  \}$ denote the family of sets of versions associated with some triple in $(s,p,o)$. We can then represent the RDF archive with a named graph for each $N \in \mathcal{N}$. However, the number of named graphs can reach $2^n$ (or the number of unique triples in $\mathcal{G}$). Another option is create a named graph for intervals~\cite{FernandezPU15}. More specifically, a triple $(s,p,o)$ is added to an interval $[i,j]$ (for $1 \leq i \leq j \leq n$) if and only if $(s,p,o) \in G_k$ for $i \leq k \leq j$ and either $i = 1$ or $(s,p,o) \notin G_{i-1}$ and $j = n$ or $(s,p,o) \notin G_{j+1}$; in simpler terms, $[i,j]$ is a maximal contiguous interval of versions in which $(s,p,o)$ appears (omitting empty intervals). The upper bound is now $n(n+1)/2$ named graphs, but triples may appear in multiple named graphs for different intervals.

In general, we expect the space to be similar to $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$; in other words, quite good. However, as the number of versions $n$ grows, $O(n^2)$ interval graphs need to be unioned in the worst-case to materialise a particular version; CB representations require processing $O(1)$ (in the case of $\mathcal{G}^n_1$ and $\mathcal{G}^1_n$) or $O(n)$ (in the case of $\mathcal{G}^n_{n-1}$ and $\mathcal{G}^{n-1}_n$) named graphs to materialise a particular version.

\paragraph{Notation:} We denote IC by $\ic$; differential, incremental, reverse-differential and reverse-incremental CB by $\cbpd$, $\cbpi$, $\cbmd$ and $\cbmi$, resp.; and interval TB by $\tb$.

@@

Given a SPARQL query $Q$ over an RDF graph, we now describe automatic rewritings of $Q$ to generate solutions for different versions. We first focus on rewritings of triple patterns and then generalise. We assume that version parameters are given via HTTP rather than extending the SPARQL syntax. For reasons of space we rather present examples in online material~\cite{online}.

@@

\paragraph{Change-Based (CB)} Recalling the observation that forwards and reverse CB representations are analogous, for brevity we define the rewriting for the forwards direction (\cbpd, \cbpi) only; the reverse direction (\cbmd, \cbmi) follows naturally.

For the differential representation, we load the base version and the positive delta into the default graph and, for each triple pattern, we subtract the negative delta which is queried using a quad pattern. Thus for $\cbpd$ we rewrite $Q$ to  \sfrom{Q'}{\{\texttt{1},\texttt{v1}\}}{\{\texttt{1v}\}}, where \texttt{1}, \texttt{v1}, \texttt{1v} indicate the names of $G_1$, $\Delta_v^1$ and $\Delta_1^v$, respectively; and $Q'$ replaces each triple pattern $(s,p,o) \in Q$ with the named graph pattern $\sminus{(s,p,o,*)}{(s,p,o,\texttt{1v})}$.

Unfortunately the incremental rewriting is more complex. A first idea would be to take the union of $G_1$ and all positive deltas $\Delta_2^1, \ldots, \Delta_v^{v-1}$ and then subtract the (union of the) negative deltas $\Delta_1^2, \ldots, \Delta_{v-1}^{v}$; unfortunately this would overlook triples that were removed from a version $1 < u < v$ but were added back in a later version $u < u' \leq v$ (and were not removed again in a version $u' < u'' \leq v$). Hence a recursive rewriting appears to be necessary. Let $Q_1 \coloneqq \sfrom{Q}{\{\texttt{1}\}}{\{\}}$; then $Q_2 \coloneqq \sfrom{Q'_1}{\{\texttt{1}\}}{\{\texttt{12},\texttt{21}\}}$, where $Q_1$ is the result of replacing each triple pattern $(s,p,o)$ in $Q_1$ by the named graph pattern $P_2 \coloneqq \sminus{\suni{(s,p,o,*)}{(s,p,o,\texttt{21})}}{(s,p,o,\texttt{12})}$. We can then apply this rewriting recursively: $Q_i \coloneqq \sfrom{Q_{i-1}'}{\{\texttt{1}\}}{\{\texttt{12},\ldots,\texttt{(i-1)i},\texttt{21},\ldots,\texttt{i(i-1)}\}}$, where $Q_{i-1}'$ replaces each named graph pattern $P_{i-1}$ appearing in $Q_{i-1}$ with the recursive pattern $P_i \coloneqq \sminus{\suni{P_{i-1}}{(s,p,o,\texttt{i(i-1)})}}{(s,p,o,\texttt{(i-1)i})}$.

This rewriting leads to complex queries. We thus optimise using additional features of SPARQL. To sketch the idea, our goal is then to make sure that each triple pattern only matches triples that appear in a base version and were not removed, or that appear in a positive delta $\Delta_{j}^{i}$ such that $1 < i < j \leq v$ and do not appear in a (later) negative delta $\Delta_{k}^{l}$ for $j \leq k < l \leq v$. In practice, for each delta $\Delta_{j}^{i}$ stored as a named graph $\texttt{ji}$, in offline processing, we index meta-data of the form $(\texttt{ji},\texttt{ver},j,\texttt{\$})$, and $(\texttt{ji},\texttt{type},\texttt{pos},\texttt{\$})$ in the case that $i < j$ or $(\texttt{ji},\texttt{type},\texttt{neg},\texttt{\$})$ in the case that $j < i$ ($\texttt{\$} \in \I$ is a reserved name for the meta-data graph). We can then check the aforementioned condition by using aggregation to find the maximum version of a negative delta less than or equals $v$ in which the triple pattern matches, then filtering the base version or any positive delta earlier than this maximum version. While the resulting query is still quite complex, we found it to be more practical than the recursive rewriting.

@@

Given a query $Q$, a control version $u$ and a target version $v$, delta-version queries give solutions in $Q(G_v) \setminus Q(G_u)$. The general strategy for rewriting is to construct a query $\sminus{Q_v}{Q_u}$ where $Q_v$ and $Q_u$ are the respective single-version queries. %While we continue to use the default graph as before for $Q_v$, we must rather use named graphs to manage the positive elements of $Q_u$ (replacing \texttt{FROM} with \texttt{FROM NAMED}/\texttt{GRAPH}). As an example, for $\ic$, we rewrite $Q$ to \sfrom{Q'}{\{\texttt{v}\}}{\{\texttt{u}\}}, where $Q'$ replaces each triple pattern $(s,p,o)$ in $Q$ with $\sminus{(s,p,o,*)}{(s,p,o,\texttt{u})}$.

@@

\paragraph{Indexing} We first look at the results of total index sizes for each representation. In Figure~\ref{fig:index} we show the index sizes (GB) for each representation, the time taken (min.) to bulk load all versions in the representation, and the time taken (min.) to update a seven-version archive with the eighth version. We see that \ic has the largest index sizes, followed by \cbpd and \cbmd, then \tb, and finally \cbpi and \cbmi. The bulk load times correlate with index size, with \ic (thus) being by far the slowest. We see a similar trend in version updates, except that \cbmd is far slower than the other alternatives (even \ic) as the entire archive must be built from scratch.
